### 2018-06-01 submission/manuscript.Rmd
updates from a final read through

# Error Introduced into 16S rRNA Gene Sequencing Results Varies by High Fidelity DNA Polymerase Used

@@

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities. It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and the  correlation between chimera prevalence and number of OTUs.

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs varied by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`).

@@

A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016]. The authors could reduce the difference between the two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016]. Yet, within this study there was no comparison made between different high fidelity DNA polymerases. Due to this gap, it is natural to extend this line of inquiry and test if biases in the number of OTUs and chimeras also are dependent on the type of high fidelity DNA polymerase. This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. We will accomplish this by examining the number of OTUs, error rate, number of sequences with an error, and chimera prevalence at varying number of cycles in five different high fidelity DNA polymerases

@@

***Human and Mock Samples:***
Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples were previously published [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR Protocol:***
The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer. The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). For Kappa and Q5, the annealing and extension time were from a previously published protocol [@taq_Gohl_2016]. For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.

@@

***Sequence Processing:***
The mothur software program was used for all sequence processing steps [@mothur_schloss_2009]. The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Analysis Workflow:***
To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis. The total number of OTUs was analyzed for both the fecal and mock community samples. For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g. 20 versus 25, 25 versus 30). Based on these observations we analyzed potential reasons for these differences. Analysis of the mock community of each polymerase for sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people. Bray-Curtis distance matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000 sequence depth. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found in the SRA at the following accession number SRP132931.

@@

***The number of OTUs generated are dependent on polymerase used:***
Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used (e.g. Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1]. Additionally, there is a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g. Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1]. There is an overall difference in the number of OTUs detected within fecal samples between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1]. Using a Tukey post-hoc test to identify which polymerase groups were different from each other at 35 cycles, no difference was found (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2]. The polymerase dependent difference in the range of the number of OTUs also was observed in the mock community samples [Figure 2]. The closest the polymerases came to the total of `r pull(ref_mock_otu_num, numOtus)` OTUs created by the mock reference 16S sequences was at 25 and 30 cycles [Figure 2]. Regardless of if fecal or mock communities were used, the same polymerases generated high and low number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S2].  In contrast to the results obtained with fecal samples, differences between polymerases for the number of OTUs created were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S1]. Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community is largely due to Kappa and Platinum differences versus the other polymerases across the different number of cycles [Table S2]. Based on these observations in fecal and mock communities, it is clear that using different polymerases result in a different total number of OTUs within a sample.

***The bacterial community is similar within polymerase and varies by number of cycles:***
Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 3]. Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`. For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 at 15 cycles. There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g. the 15 cycle versus 20 cycle group compared to the 30 cycle to 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A]. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05). The mock community has similar trends to the observations reported for the fecal bacterial community but none were significant `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B]. Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.

***Sequence error varies by polymerase and is consistent across the number of cycles used:***
The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3]. The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 and Table S4]. The total sequences with at least one error is also polymerase dependent with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4]. These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.

@@

Standardization of the 16S rRNA gene sequencing workflow may partially solve this problem by introducing a consistent bias to all samples. However, as mentioned in the introduction, many of the choices made to standardize the workflow will result in missing important members of the microbial community. Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis. All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that wants to be answered. Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occurring.

@@

Our observations show that errors dependent on either high fidelity DNA polymerases used or the number of cycles chosen can change diversity metrics used in 16S rRNA gene sequencing results. Care should be taken when choosing a polymerase for 16S rRNA gene surveys because they can influence diversity-based metrics. By knowing the inherent bias associated with different polymerases it allows for more informed choices to be made on how to reduce 16S rRNA gene sequencing-based error.

@@

The authors would like to thank all the study participants in ERIN whose samples were utilized. We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan. Additional thanks to members of the Schloss lab and Dr. Marcy Balunas for reading earlier drafts of the manuscript and providing helpful critiques. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

>>>>

# Error Introduced into 16S rRNA Gene Sequencing Varies by High Fidelity DNA Polymerase Used

@@

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities. It also assessed polymerase and number of cycle dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and the  correlation between chimera prevalence and number of OTUs.

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases (eg. at 35 cycles Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs varied by HiFi DNA polymerase used at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`).

@@

A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016]. The authors could reduce the difference between the two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016]. Yet, within this study there was no comparison made between different high fidelity DNA polymerases. Due to this gap, it is natural to extend this line of inquiry and test if biases in the number of OTUs and chimeras also are dependent on the type of high fidelity DNA polymerase. This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. We will accomplish this by examining the number of OTUs, Bray-Curtis index, error rate, number of sequences with an error, and chimera prevalence at varying number of cycles in five different high fidelity DNA polymerases

@@

***Human and mock samples:***
Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples were previously published [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR protocol:***
The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer. The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). For Kappa and Q5, the annealing and extension time also were from a previously published protocol [@taq_Gohl_2016]. For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.

@@

***Sequence processing:***
The mothur software program was used for all sequence processing steps [@mothur_schloss_2009]. The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P. aeruginosa*, *E. coli*, *S. enterica*, *L. fermentum*, *E. faecalis*, *S. aureus*, *L. monocytogenes*, and *B. subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Analysis workflow:***
To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis. The total number of OTUs was analyzed for both the fecal and mock community samples. For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g. 20 versus 25, 25 versus 30). Based on these observations we analyzed potential reasons for these differences using the mock community samples. Analysis of the mock community of each polymerase for sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.

***Statistical analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the comparison in fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people. Bray-Curtis distance matrices were generated using mothur. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable, correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Reproducible methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.

@@

***The number of OTUs generated are dependent on polymerase used:***
Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used (e.g. Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1]. Additionally, there is a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g. Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1]. There is an overall difference in the number of OTUs detected within fecal samples between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1], but a Tukey post-hoc test failed to identify which specific polymerase groups were different (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2]. The polymerase dependent difference in the range of the number of OTUs also was observed in the mock community samples [Figure 2]. The closest the polymerases came to the total of `r pull(ref_mock_otu_num, numOtus)` OTUs created by the mock reference 16S sequences was at 25 and 30 cycles [Figure 2]. Regardless of if fecal or mock communities were used, the same polymerases generated high and low number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S2].  In contrast to the results obtained with fecal samples, differences between polymerases for the number of OTUs created were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S1]. Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community was largely due to Kappa and Platinum versus the other polymerases across the number of cycles used [Table S2]. Based on these observations in fecal and mock communities, it is clear that using different polymerases result in a different total number of OTUs within a sample.

***The bacterial communities generated by the polymerases were similar but varied by number of cycles:***
Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 3]. Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`). For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles. There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g. the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A]. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05). Similar trends were observed in the mock community but none were significant (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B]. Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.

***Sequence error varies by polymerase and is consistent across the number of cycles used:***
The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3]. The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 & Table S4]. The total sequences with at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4]. These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.

@@

Standardization of the 16S rRNA gene sequencing workflow may partially solve this problem by introducing a consistent bias to all samples. However, as mentioned in the introduction, many of the choices made to standardize the workflow will result in missing important members of the microbial community. Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis. All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that want to be answered. Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occurring.

@@

Our observations show that errors dependent on either high fidelity DNA polymerases used or the number of cycles chosen can change diversity metrics used in 16S rRNA gene sequencing. Care should be taken when choosing a polymerase for 16S rRNA gene surveys because they can influence diversity-based metrics. By knowing the inherent bias associated with different polymerases it allows for more informed choices to be made on how to reduce 16S rRNA gene sequencing-based error.

@@

The authors would like to thank all the study participants in ERIN whose samples were utilized. We also would like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan. Additional thanks to members of the Schloss lab and Dr. Marcy Balunas for reading earlier drafts of the manuscript and providing helpful critiques. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.
